Looking for a Signal in the Noise: Revisiting Obesity and the Microbiome

### 2016-05-25 submission/Sze_Obesity_PeerJ_2016.Rmd
abstract text modifications

alpha_tests.summary <- read.table("../data/process/alpha_tests.summary", header=T)
alpha_composite.summary <- read.table("../data/process/alpha_composite.summary")
alpha_power.obs <- read.table("../data/process/alpha_power.observed", 
                              header = T, row.names = 1)
alpha_power.predicted <- read.table("../data/process/alpha_power.predicted", 
                                    header = T)
alpha.data.ALL <- read.table("../data/process/alpha.data", header = T)
alpha_effectSize <- read.table("../data/process/alpha_EffectSize.summary")

@@

@@

**Methods:** To identify datasets that we included in the meta analysis we followed the Preferred Reporting Items for Systematic Reviews and Meta-Analyses (PRISMA) guidelines we winnowed an initial set of 196 studies that tested the microbiome-obesity association hypothesis down to 8 studies that fit our criteria. We applied three analysis pipelines to the data from each study. First, we tested the association between community structure and diversity and the Bacteroides to Firmicutes ratio (B/F) with the subjects' body mass index (BMI) treated as a continuous or categorical value. We also used the random forest machine learning algorithm to identify community features that would allow us to to predict whether a person was obese based on either genus or OTU level. Second, we pooled the relative risk (RR) ratio of being obese based on characteristics of the microbiome for each study and tested for an association between RR and community diversity and B/F. Finally, we normalized the community diversity and B/F values within each dataset and pooled the data to test for an association with obesity.

@@

**Conclusion:** We show that by pooling the results of many individual studies we can test the generalizability of the microbiome. Our meta-analysis provides support for the null hypothesis that there is no difference between non-obese and obese individuals in Shannon diversity and B/F ratio, that the effect sizes are small, and that a very large number of indivdiuals would have to be tested to reach sufficient study power.

@@

**Table 1. Summary Demographics of Individuals used in the Meta-analysis.**
```{r Table_1, echo=FALSE, message=FALSE, warning=FALSE}

table1 <- read.table("../results/tables/table1.demographics", header=T)

>>>>

alpha_tests.summary <- read.table("../data/process/alpha_tests.summary", header=T)
alpha_composite.summary <- read.table("../data/process/alpha_composite.summary")
alpha_power.obs <- read.table("../data/process/alpha_power.observed", 
                              header = T, row.names = 1)
alpha_power.predicted <- read.table("../data/process/alpha_power.predicted", 
                                    header = T)
alpha.data.ALL <- read.table("../data/process/alpha.data", header = T)

@@

#Using a quick for loop to generate the necessary effect size information
variableTested <- seq(1:6)
variables <- colnames(alpha_power.obs)[1:6]
alpha_effectSize <- matrix(nrow=length(rownames(beta_tests.summary)), ncol=length(variableTested))
studies <- rownames(beta_tests.summary)

for(i in variableTested){
  tempdata <- c()
  for(j in studies){
  tempdata <- rbind(tempdata, abs((mean(alpha.data.ALL[which(alpha.data.ALL$dataset == j & alpha.data.ALL$obese == FALSE), variables[i]]) - mean(alpha.data.ALL[which(alpha.data.ALL$dataset == j & alpha.data.ALL$obese == TRUE), variables[i]]))/mean(alpha.data.ALL[, variables[i]]) * 100))
  }
  alpha_effectSize[, i] <- tempdata
}

colnames(alpha_effectSize) <- variables
rownames(alpha_effectSize) <- studies

@@

**Methods:** To identify datasets that we included in the meta analysis we followed the Preferred Reporting Items for Systematic Reviews and Meta-Analyses (PRISMA) guidelines we winnowed an initial set of 196 studies that tested the microbiome-obesity association hypothesis down to 10 studies that fit our criteria. We applied three analysis pipelines to the data from each study. First, we tested the association between community structure and diversity and the Bacteroides to Firmicutes ratio (B/F) with the subjects' body mass index (BMI) treated as a continuous or categorical value. We also used the random forest machine learning algorithm to identify community features that would allow us to to classify whether a person was obese based on either genus or OTU level. Second, we pooled the relative risk (RR) ratio of being obese based on characteristics of the microbiome for each study and tested for an association between RR and community diversity and B/F. Finally, we utilized linear mixed-effect models to correct for variation by study and analyzed the pooled community diversity and B/F values within each dataset to test for an association with obesity.

@@

**Conclusion:** We show that by pooling the results of many individual studies we can test the generalizability of the microbiome and its effect on obesity. Our meta-analysis provides support for the observation that there is a difference between non-obese and obese individuals in Shannon diversity but that the effect sizes are small, and that a very large number of indivdiuals would have to be tested to reach sufficient study power.

@@

**Table 1. Summary Demographics of Individuals used in the Meta-analysis.**
```{r Table_1, echo=FALSE, message=FALSE, warning=FALSE, eval=FALSE}

Study <- as.character(rownames(beta_tests.summary))

Age <- as.character(demographics$meanAge)
AgeSD <- as.character(format(round(
  demographics$SDAge, 2), nsmall = 2))
AgeTable <- paste(Age," (",AgeSD,")", sep="")

Sex <- paste(demographics$females, "|", demographics$males)

Ancestry <- as.character(format(round(
  demographics$ancestry*100, 2), nsmall = 2))

ObeseTotal <- ShannonRRTable$tposH + ShannonRRTable$cposH
ContTotal <- ShannonRRTable$tnegH + ShannonRRTable$cnegH

ObBreakdown <- paste(ContTotal, " | ", ObeseTotal, sep="")

BMI <- paste(demographics$meanBMI, " (", demographics$SDBMI, ")", sep="")

Min <- as.character(demographics$minBMI)

Max <- as.character(demographics$maxBMI)

table1 <- cbind(Study, AgeTable, Sex, Ancestry, ObBreakdown, BMI, Min, Max)
table1[table1=="NA" | table1=="NA | NA" | table1=="NA (NA)" |
         table1=="      NA"] <- "Not Available"
